---
title: "Predicting Body Fat Percentage using Anthropometric Measurements"
name: "Matthew Yoon"
output: html_document
date: "2025-06-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 2.1 Title

title: "Predicting Body Fat Percentage using Anthropometric Measurements"
name: "Matthew Yoon"

## 2.2 Abstract

According to the given data set from the SPSS On-Line Training Workshop, the data is divided into four categories: "amount of body fat", "triceps skinfold thickness", "thigh circumference", and "midarm circumference". According to my findings, it is apparent that among these Anthropometric measurements only "triceps skinfold thickness" and "thigh circumference" are the significant predictors of body fat percentage. The type of linear model used to come to these results were multiple linear regressions. The final model equation determined from this analysis was Bodyfat = -32.32719 + 0.83303 * Triceps + 0.52401 * Thigh + 0.02638 * Midarm. The model's goodness of fit or the adjust r-squared is 0.9985 which indicates a very strong fit due to the high percentage in variance in the dependent variables.  

### 2.3 Introduction

Understanding one's body fat is crucial for optimizing one's health. Body fat percentage can in fact impact overall well-being and disease risk. Although fat serves vital functions like insulation, excess or deficient levels can lead to various health risks. For example, high body fat percentages are linked to increased risks of Type 2 diabetes, high blood pressure, and cardiovascular diseases. Especially concerning visceral fat which is the fat stored around organs, it can lead to serious health risks like heart disease and stroke. Extremely low body fat percentages can potentially lead to osteoporosis and increased risk of injury. Rather than relying solely body weight or BMI, body fat percentage can provide a more accurate picture of one's health. There are numerous ways to measure one's body fat including skinfold thickness and underwater weighing however, anthropometric measurements are seen as cost-effective alternatives and include numerous advantages including its non-invasive and easily obtainable nature, helpful for tracking change over time, and can be used to assess the overall adiposity. Certain limitations however include that it may not be as accurate as methods such as a Dual-Energy X-ray absorptiometry and anthropometric prediction equations developed from specific sample populations may not be reliable for others. Potential research questions that this analysis aims to answer include "Can a linear model effectively predict body fat percentage using the mentioned category predictors?" and "Which of the anthropometric measurements are the most significant predictors of body fat percentage?" 

### Description of Data

```{r}
library(ggplot2)
library(dplyr)

bodyfat_data <- data.frame(
  Bodyfat = c(19.5, 24.7, 30.7, 29.8, 19.1, 25.6, 31.4, 27.9, 22.1, 25.5, 31.1, 30.4, 18.7, 19.7, 14.6, 29.5, 27.7, 30.2, 22.7, 25.2),
  Triceps = c(43.1, 49.8, 51.9, 54.3, 42.2, 53.9, 58.5, 52.1, 49.9, 53.5, 56.6, 56.7, 46.5, 44.2, 42.7, 54.4, 55.3, 58.6, 48.2, 51.0),
  Thigh = c(29.1, 28.2, 37.0, 31.1, 30.9, 23.7, 27.6, 30.6, 23.2, 24.8, 30.0, 28.3, 23.0, 28.6, 21.3, 30.1, 25.7, 24.6, 27.1, 27.5),
  Midarm = c(11.9, 22.8, 18.7, 20.1, 12.9, 21.7, 27.1, 25.4, 21.3, 19.3, 25.4, 27.2, 11.7, 17.8, 12.8, 23.9, 22.6, 25.4, 14.8, 21.1)
)

summary(bodyfat_data)

hist(bodyfat_data$Bodyfat, main = "Distribution of Body Fat Percentage", xlab = "Body Fat Percentage")
hist(bodyfat_data$Triceps, main = "Distribution of Body Fat Percentage", xlab = "Body Fat Percentage")
hist(bodyfat_data$Thigh, main = "Distribution of Thigh Circumference", xlab = "Thigh Circumference")
hist(bodyfat_data$Midarm, main = "Distribution of Midarm Circumference", xlab = "Midarm Circumference")

ggplot(bodyfat_data, aes(x = Triceps, y = Bodyfat)) + geom_point() + labs(title = "Body Fat vs. Triceps Skinfold Thickness", xlab = "Triceps Skinfold Thickness", ylab = "Body Fat Percentage")

ggplot(bodyfat_data, aes(x = Thigh, y = Bodyfat)) + geom_point() + labs(title = "Body Fat vs. Thigh Circumference", xlab = "Thigh Circumference", ylab = "Body Fat Percentage")

ggplot(bodyfat_data, aes(x = Midarm, y = Bodyfat)) + geom_point() + labs(title = "Body Fat vs. Midarm Circumference", xlab = "Midarm Circumference", ylab = "Body Fat Percentage")
```

### 2.5 Methods and Results

```{r}
linear_model <- lm(Bodyfat ~ Triceps + Thigh + Midarm, data = bodyfat_data)

summary(linear_model)

plot(linear_model, which = 1)

plot(linear_model, which = 2)

plot(linear_model, which = 3)

plot(linear_model, which = 5)
```
According to these results, the residuals are the differences between the observed values and the values predicted by the model. In an ideal world, these values should be balanced around zero which means the model fits the data well without a far estimation. The median being -0.00878 indicates that the residuals are centered around zero. The minimum and maximum residuals are relatively close in absolute value and the first and third quartiles have similar magnitudes meaning there is a reasonable distribution around zero. According to the coefficients, the intercept -32.32719 represents the model's baseline prediction when the predictors are zero. For triceps, for every one-unit increase in the triceps variable, the dependent variable is estimated to increase by roughly 0.833 and this notion applies to the other factors. Because the Midarm has a p-value of 0.17 which is greater than the typical 0.05 threshold, this coefficient is not statistically significant at the 0.05 level meaning it cant be concluded that it has a nonzero effect on the dependent variable. Based on the model fit statistics, the multiple R-squared value of 0.9988 indicates that about 99.88% of the variance in the dependent variable can be elaborated by the independent variables which suggests a strong linear relationship. The adjusted R-squared value of 0.9985 is very close to the R-squared value which indicates that the added predictors are affecting the model significantly. The F-statistic of 4263 on 3 and 16 DF and p-value of <2.2e-16 shows that with a large F-statistic and very small p-value indicates that the model is statistically significant meaning at least one of the independent variables is related to the dependent variable. This suggests that the regression model is a better fit of the data as opposed to a model with no independent variables. 

### 2.6 Conclusion

The Key Findings of this research include the Model Fit which appears to be a good fit for the data. The high R-squared value of 0.9988 shows that the independent variables explain 99.88 percent of variability in body fat and the adjusted R-squared of 0.9985 is also high and close to the R-squared value which means that the model included variables that contributed meaningfully to bodyfat variability. The F-statistic of 4263 with a low p-value justifies the significance of the model and its independent variables. The variable "Triceps" and "Thighs" are statistically significant predictors of bodyfat indicated by the low p-values. Given the positive coefficients, as triceps and thigh measurements increase, bodyfat tends to increase. Thus Midarm measurements are nit reliably associated with bodyfat changes. Potential areas for research include researching why the Midarm variable was not significant which could involve learning more about potential multicollinearity variables and further data collection by considering a larger and more diverse dataset to ensure the findings can be applied to a broader population. 

